Type Ia Supernova Model

### 2015-07-02 doc/model.tex
version given locally

\section{Model}
We infer a Hubble diagram based on measurements of discovered transients
classified as Type~Ia supernovae (SNe~Ia).  The inferred Hubble diagram can be
expressed as
\begin{equation}
P(\vec{\mu},\vec{z}| \vec{\hat{D}}=1, \vec{\hat{ADU}}, \vec{\hat{T}}_S=\text{SN~Ia},\vec{\hat{z}}_S,
\vec{\hat{\theta}}_S,\vec{\hat{z}}_H,\vec{\hat{\theta}}_H),
\label{hd:eqn}
\end{equation}
where the inferred quantities are the set of distance moduli ($\vec{\mu}$)
and redshifts ($\vec{z}$), and 
the measurements include: $\hat{D}$ the boolean that indicates that
the transient was discovered; $\hat{ADU}$ the transient photometry;
spectroscopically derived transient
typing, redshift and parameters $\hat{T}_S$, $\hat{z}_S$, $\hat{\theta}_S$;
and the redshift and parameters of the associated host galaxy $\hat{z}_{H}$, $\hat{\theta}_H$.
The conditions $\vec{\hat{D}}=1$, $\vec{\hat{T}}_S,=\text{SN~Ia}$ hold as only transients
discovered and classified as SNe~Ia are modelled.  We use the convention of using carets $\hat{}$
to  label measurements.

A sketch of a model explaining the data is depicted through the Probabilistic Graphical Model
(PGM)
shown in Figure~\ref{pgm:fig}.
Although only transients that are detected and classified as
SN~Ia are modeled, the model accommodates the possibility that they could not have been found
or not classified as SN~Ia.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=7in]{/Users/akim/project/abc/results/hdpgm.eps} 
   \caption{Probabilistic Graphical Model for the SN~Ia analysis.
   The bottom right corner containing the connection between observed light curves
   $\hat{f}$, transient spectral information $\hat{z}_S$, $\hat{\theta}_S$, and
   host information $\hat{z}_H$, $\hat{\theta}_H$ with the determined type, redshift, and distance
   modulus $\hat{T}$, $\hat{z}$, and $\hat{\mu}$ is the purview of
   the traditional light-curve and Hubble diagram fitters.
   \label{pgm:fig}}
\end{figure}

One feature to note is that $\mu$ is a model parameter; most analyses instead
directly have cosmological parameters $\{\Omega_M, \Omega_\Lambda, w_0, w_a\}$.
A motivation is to provide results that can be applied to test arbitrary cosmological models:
indeed the model-dependence of the $\mu$'s reported for high-$z$ supernovae 
\citep[e.g.][]{2012ApJ...746...85S} is a source of confusion to the community.
An example of a non-parametric model for $\mu$ that is independent of cosmological
parameters is given in the Appendix of
\citet{2013ApJ...764..116W}.

@@

Discovered transients typed as SNe~Ia are modeled.  Each has
a cosmological redshift and corresponding distance modulus; these $z$-$\mu$ pairs are the  parameters that
constitute the Hubble diagram.  For the moment, the coordinates RA/Dec are considered
precisely determined from the images.

Each transient indexed by $i$ is of a certain type $T$ and comes from a  host galaxy of type $G$.  
(``Host'' includes the host as well as  all galaxies that project onto the coordinate of
the transient.)  The
transients (galaxies) are modeled and described by global parameters $\theta_T$ ($\theta_G$),
and  parameters specific to the transient $\theta_{Ti}$ ($\theta_{Gi}$).  For example, the SALT2 SN~Ia
model has individual $c$ and $x_1$ parameters that would fill $\theta_{Ti}$,
and global $\alpha$, $\beta$, and intrinsic dispersion  parameters  common to all supernovae
 and  underlying  $c$ and $x_1$ distributions that would fill $\theta_T$.
The Type model includes  information on the
source and line-of-sight effects, which specify the effective SED. 
This SED, redshift, and  distance modulus determine the photon flux incident above the Earth's
atmosphere, $n$ and $n_g$.

@@

For discovered transients, there may be spectra of the object itself
from which typing, redshift and spectral parameters $\hat{T}_S$, $\hat{z}_S$, $\hat{\theta}_S$ are derived.
From amongst the detected galaxies $\hat{\mathit{Gals.}}$  a host is associated with the transient
based on its RA/Dec, its
host redshift (spectroscopic or photometric) and parameters are $\hat{z}_{H}$, $\hat{\theta}_H$.

@@

The observables are:
\begin{description}
\item[$\hat{\mathit{Gals}}$] Detected galaxies from which a host is associated with a transient.
Depends on the model for galaxies $\theta_G$.
\item[$\hat{D}$] Transient detection.  It depends on
the realized signal $\hat{\mathit{ADU}_V}$ and its noise, mostly through the signal-to-noise
$\mathit{SNR}$.
\item[$\hat{\mathit{ADU}}$]  Photometry of discovered transients.  Exactly the value
of the  realized signal $\mathit{ADU}_V$ and discovery $\hat{D}$.
\item[$\hat{z}_H$, $\hat{\theta}_H$] Host galaxy redshift and properties from spectroscopic
and photometric data, based on association
of the host from the pool of detected galaxies $\hat{\mathit{Gals}}$ and the transient RA/Dec.
\item[$\hat{T}_S$, $\hat{z}_S$, $\hat{\theta}_S$] Transient redshift and properties from
spectroscopic data. I skip the details of the actual measurement  and make a direct association
with the fluxes of the transient and underlying galaxy $n$, $n_g$.
\item[$\hat{Z}$] Calibration zeropoints.  Depends on the calibration process
and the underlying transmission functions $\phi(\lambda)$.  This may be a zeropoint
but most probably will be calibrated transmission functions.
\end{description}

The following are elements of the model.
\begin{description}
\item[$\mathit{ADU}_V$]  Realized counts
of the transient.  Depends
on the expected counts of the transient and host $\mathit{ADU}$, $\mathit{ADU}_g$.
This is distinct from $\hat{\mathit{ADU}}$, which only has photometry of detected transients.
\item[$\mathit{ADU}$, $\mathit{ADU}_{g}$]  Expected counts
of the transient and background galaxy respectively.  Depends
on the incident fluxes $n$, $n_g$ and the instrumental
transmission $\phi(\lambda)$.
\item[$\phi(\lambda)$] DES transmission functions.
\item[$n$, $n_g$] The incident fluxes at Earth of the transient and background
galaxy.  Depends on  $z$ and $\mu$, and the SEDs of the Type and Host as
specified by $\theta_{Ti}$, $\theta_{Gi}$,$\theta_{T}$, $\theta_{G}$.
\item[$\theta_{Ti}$, $\theta_{Gi}$] Parameters that describe the properties of the individual transient
and host respectively.
\item[$\theta_{T}$, $\theta_{G}$] Parameters that describe the global properties of the transient
and host models.
\item[$T$,$G$]  The type of transient and projected galaxies (referred to as host), which
specifies which model to use. The realization depends on the redshift $z$.
\item[$\vec{\mu}$, $\vec{z}$]  The true distance moduli and cosmological redshifts under
consideration.  Each redshift has a unique distance modulus.
\item[$\mathbf{RA/Dec}$] The coordinate of the transient.  It will be the actual coordinate of the found discovered object.
\end{description}

@@

\subsection{Type}
\label{type:sec}
Given that our transient sample may not consist purely of SNe~Ia, the model
has a term of the form $P(T, G | z,\mu)$, which should account for all types of
objects that could potentially be mistaken for Type~Ia:
the rates and association with host galaxies are described for
the full population of potential interlopers.
%Note the condition $\hat{T}=\text{SN~Ia}$ means that non-Ia transients unlikely to
%be identified as SN~Ia can be described with poorer accuracy.
This term is not currently well known, particularly at the highest redshifts probed by DES.
It could be constructed through modeling, although then the uncertainties 
of that modeling must in turn be quantified.

I posit that a concise way of addressing this term is to only model $T=\text{SN~Ia}$,
and experimentally determine $P(\text{non-Ia})$, 
$\left. \langle \mu-\mu_{model} \rangle \right|_{non-Ia}$, and its uncertainty for the
sample of non-Ia's that are discovered and misclassified as SN~Ia.
This measurement is achieved through spectroscopic
typing of a subset drawn from the populations used in the Hubble diagram.
Other supernova surveys
with spectroscopic follow-up (e.g.\ SNLS) that replicate the DES population
can be used, for example on magnitude-limited samples with
complete  spectroscopic follow-up. 
The weighted bias would be added as a correction to the Hubble diagram under the assumption that all objects are SNe~Ia.

To minimize the contribution of non-Ia contamination we strive for
an excellent classifier so that $P(non-Ia) \rightarrow 0$.

\subsection{Host Matching}
\sloppy
The host-galaxy properties depends on the galaxy catalog and the transient
coordinates $P(\hat{z}_H, \hat{\theta}_H | \hat{\mathit{Gals.}}, \text{RA/Dec})$.  With a
spectrum of the transient and the underlying background, identification of the host galaxy
is usually straightforward.
Otherwise, projections or ambiguous cases can result in the misidentification of
the host.

A similar strategy can be taken as with non-Ia contamination in \S\ref{type:sec}:
analyze the supernova assuming a correct host match, and
experimentally determine $P(\text{mismatch} )$ and
$\left. \langle \hat{\mu}-\mu(z)\rangle \right|_{\text{mismatch}}$, $\left. \langle \hat{z}-z \rangle \right|_{\text{mismatch}}$ to correct bias in the Hubble diagram. 
We strive to have $P(\text{mismatch}) \rightarrow 0$.

\subsection{Type~Ia Supernova Model}
\label{snmodel:sec}
The model contains a piece $P(\hat{\mathit{ADU}}_{i} \hat{z}_{Si}| \mu_i,z_i,\theta_T, \theta_{Ti}, ,\phi(\lambda))$, which
is the piece that describes underlying SN~Ia properties and light-curve fits.

Need something more general?
$P(\hat{D}_i,\hat{\mathit{ADU}}_{i}, \hat{T}_{Si}, \hat{z}_{Si}, \hat{\theta}_{Si}, \hat{z}_{Hi}, \hat{\theta}_{Hi}
 | \mu_i,z_i,T_i,\theta_T,\theta_{Ti},\phi(\lambda))$
\subsection{Detection Efficiency}
There term $P(\hat{D}|\mathit{ADU}_V)$.  For all high-redshift supernova searches
I know of, this depends on the signal-to-noise of the data, and very weakly on the
structure of the underlying galaxy.  Our detection is now based on machine learning,
we should confirm that detection dependencies do not depend on our model parameters
(except for RA/Dec).

\section{Systematic Tests}
The Hubble diagrams of subsets of data quantify systematic uncertainty.  
Subsets that may be expected to show evidence for systematics include:
spectroscopically typed SNe~Ia; transients in deep (3) versus shallow (1/2) DES fields;
transients with NIR data; splits based on host properties.

>>>>

\section{Model}
We infer a Hubble diagram based on measurements of discovered transients
classified as Type~Ia supernovae (SNe~Ia).  The inferred Hubble diagram can be
expressed as
\begin{equation}
P(\vec{\mu},\vec{z} |  \vec{\hat{D}}=1, \vec{\hat{T}}=\text{SN~Ia}, \vec{\hat{ADU}}, \vec{\hat{T}}_S,\vec{\hat{z}}_S,
\vec{\hat{\theta}}_S,\vec{\hat{z}}_H,\vec{\hat{\theta}}_H),
\label{hd:eqn}
\end{equation}
where the inferred quantities are the set of distance moduli ($\vec{\mu}$)
and redshifts ($\vec{z}$), and 
the measurements include: $\hat{D}$ the boolean that indicates that
the transient was discovered; $\hat{ADU}$ the transient photometry;
spectroscopically derived transient
typing, redshift and parameters $\hat{T}_S$, $\hat{z}_S$, $\hat{\theta}_S$;
and the redshift and parameters of the associated host galaxy $\hat{z}_{H}$, $\hat{\theta}_H$.
A classifier takes these data and infers types $\vec{\hat{T}}$.
The conditions $\vec{\hat{D}}=1$, $\vec{\hat{T}}=\text{SN~Ia}$ hold as only transients
discovered and classified as SNe~Ia are modeled.  We use the convention of using carets $\hat{}$
to  label measurements.

An alternative would have been to
model measurements of all discovered transients, not just those typed as SN~Ia. 
The non-Ia transient population has significant model uncertainty and the information they impart on the expansion history is weak:  I anticipate it is cleaner to limit ourselves to
modeling discoveries we think are SN~Ia.  This restriction does not obviate the need
to consider non-Ia contamination in the sample.

Here $\mu$ is modeled; most analyses instead
directly have cosmological parameters $\{\Omega_M, \Omega_\Lambda, w_0, w_a\}$.
A motivation is to provide results that can be applied to test arbitrary cosmological models:
indeed the model-dependence of the $\mu$'s reported for high-$z$ supernovae 
\citep[e.g.][]{2012ApJ...746...85S} is a source of confusion to the community.
An example of a non-parametric model for $\mu$ that is independent of cosmological
parameters is given in the Appendix of
\citet{2013ApJ...764..116W}.

DES plans a ``photometry''-only analysis, in that a significant subset of transients used
in the Hubble diagram will not observed spectroscopically while active.  Therefore,
our model must account for transient mistyping and host-galaxy mismatches.  To obtain
our targeted
accuracy we need to account for covariances that occur due to photometric
calibration and Malmquist bias.  Our model will contain elements that may not
have been included in previous analyses.

A sketch of a model explaining the data is depicted through the Probabilistic Graphical Model
(PGM)
shown in Figure~\ref{pgm:fig}.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=7in]{/Users/akim/project/abc/results/hdpgm.eps} 
   \caption{Probabilistic Graphical Model for the SN~Ia analysis.  The forward modeling
   flow goes from left to right.  Starting from the position and distance
   $\mu$, $z$, $\text{RA/Dec}$ the transient and host types $T$, $G$ are realized. 
   The transient- and
   galaxy-model parameters $\theta_T$, $\theta_{Ti}$, $\theta_G$, $\theta_{Gi}$ along
   with redshift and distance modulus determine the incoming photon flux $n$, $n_g$.
   The instrumental transmission function $\phi$ gives the expected and realized
   counts $\mathit{ADU}$, $\mathit{ADU}_g$, $\mathit{ADU}_V$.  The calibration measurements
   are $Z$.  The measured galaxy catalog is $\hat{\mathit{Gals.}}$  Only detected
   supernovae $\hat{D}$ have subsequent light curve ($\hat{ADU}$), spectral
   $\hat{X}_S$, and associated host galaxy $\hat{X}_H$ measurements.  A supernova
   classifier takes measurements and infers a type $\hat{T}$, which is used for sample selection.
   \label{pgm:fig}}
\end{figure}

@@

Discovered transients typed as SNe~Ia are modeled.  Each has
a cosmological redshift and corresponding distance modulus; these $z$-$\mu$ pairs are the  parameters that
constitute the Hubble diagram.  For the moment, the coordinates RA/Dec are considered
precisely determined from imaging data.

Each transient indexed by $i$ is of a certain type $T$ and comes from a  host galaxy of type $G$.  
(``Host'' includes the host as well as  all galaxies that project onto the coordinate of
the transient.)  The
transients (galaxies) are modeled and described by global parameters $\theta_T$ ($\theta_G$),
and  parameters specific to the transient $\theta_{Ti}$ ($\theta_{Gi}$).  For example, the SALT2 SN~Ia
model has individual $c$ and $x_1$ parameters that would constitute $\theta_{Ti}$,
and global $\alpha$, $\beta$, and intrinsic dispersion  parameters  common to all supernovae
 and  underlying  $c$ and $x_1$ distributions that would constitute $\theta_T$.
The Type model includes  information on the
source and line-of-sight effects, which specify the effective SED. 
This SED, redshift, and  distance modulus determine the photon flux incident above the Earth's
atmosphere, $n$ and $n_g$.

@@

For discovered transients, there may be spectra of the object itself
from which typing, redshift and spectral parameters $\hat{T}_S$, $\hat{z}_S$, $\hat{\theta}_S$ are derived.
From amongst the detected galaxies $\hat{\mathit{Gals.}}$  a host is associated with the transient
based on its RA/Dec; the host information is summarized as the
host redshift (spectroscopic or photometric) and parameters $\hat{z}_{H}$, $\hat{\theta}_H$.

@@

The observables are:
\begin{description}
\item[$\hat{\mathit{Gals}}$] Detected galaxies from which a transient's host is identified.
Depends on the model for galaxies $\theta_G$.
\item[$\hat{D}$] Transient detection.  It depends on
the realized signal $\hat{\mathit{ADU}_V}$ and its noise, mostly through the signal-to-noise
$\mathit{SNR}$.
\item[$\hat{\mathit{ADU}}$]  Photometry of discovered transients.  Its inclusion
depends on discovery $\hat{D}$ and has the value of the realized signal $\mathit{ADU}_V$.
\item[$\hat{z}_H$, $\hat{\theta}_H$] Host galaxy redshift and properties from spectroscopic
and photometric data, based on association
of the host from the pool of detected galaxies $\hat{\mathit{Gals}}$ and the transient RA/Dec.
\item[$\hat{T}_S$, $\hat{z}_S$, $\hat{\theta}_S$] Transient redshift and properties from
spectroscopic data. I skip the details of the actual measurement  and make a direct association
with the fluxes of the transient and underlying galaxy $n$, $n_g$.
\item[$\hat{Z}$] Calibration zeropoints.  Depends on the calibration process
and the underlying transmission functions $\phi(\lambda)$.  Specifies
the transmission functions.
\end{description}

The following are elements that are to be modeled, including ``nuisance'' parameters.
\begin{description}
\item[$\mathit{ADU}_V$]  Realized counts
of the transient.  Depends
on the expected counts of the transient and host $\mathit{ADU}$, $\mathit{ADU}_g$.
Although all transients will have realized counts but not all will be measured, hence the distinction
from $\hat{\mathit{ADU}}$.
\item[$\mathit{ADU}$, $\mathit{ADU}_{g}$]  Expected counts
of the transient and background galaxy respectively.  Depends
on the incident fluxes $n$, $n_g$ and the instrumental
transmission $\phi(\lambda)$.
\item[$\phi(\lambda)$] DES transmission functions.
\item[$n$, $n_g$] The incident fluxes at Earth of the transient and background
galaxy.  Depends on  $z$ and $\mu$, and the SEDs of the Type and Host as
specified by $\theta_{Ti}$, $\theta_{Gi}$, $\theta_{T}$, $\theta_{G}$.
\item[$\theta_{Ti}$, $\theta_{Gi}$] Parameters that describe the properties of the individual transient
and host respectively.
\item[$\theta_{T}$, $\theta_{G}$] Parameters that describe the global properties of the transient
and host models.
\item[$T$, $G$]  The type of transient and projected galaxies (referred to as host), which
specifies which model to use. The realization depends on the redshift $z$.
\item[$\vec{\mu}$, $\vec{z}$]  The true distance moduli and cosmological redshifts under
consideration.  Each redshift has a unique distance modulus.
\item[$\mathbf{RA/Dec}$] The coordinate of the transient.  For us, it is sufficient to directly
assign the coordinate measured
from the images.
\end{description}

@@

\subsection{Type}
\label{type:sec}
Since our transient sample may not consist purely of SNe~Ia, the model
has a term of the form $P(T, G | z,\mu)$, which should account for all types of
objects that could potentially be mistaken for Type~Ia:
the rates and association with host galaxies are described for
the full population of potential interlopers.
This term is not currently well known, particularly at the highest redshifts probed by DES.

An idea is to consider two types: SN~Ia and non-SN~Ia, where the latter's
intrinsic distributions have loose priors.  Transients that have been spectroscopically
classified as non-Ia will provide the strongest leverage in constraining the
contamination: this subset must be unbiased or properly weighted
relative to the underlying population that constitutes the Hubble diagram.
Running the model on a pure sample of SNe~Ia can have benefits.
As discussed in \S\ref{systematics:sec}, comparing these results with
those from the full sample provides
a test of systematics.  Alternatively, the non-Ia model may be better constrained
with the pure sample.

To minimize the contribution of non-Ia contamination we strive for
a pure sample selection so that $P(T=\text{non-Ia}| \mu) \rightarrow 0$.

\subsection{Host Matching}
\sloppy
The host-galaxy properties depend on the galaxy catalog and the transient
coordinates $P(\hat{z}_H, \hat{\theta}_H | \hat{\mathit{Gals.}}, \text{RA/Dec})$.  With a
spectrum of the transient and the underlying background, identification of the host galaxy
is usually straightforward.
Otherwise, projections or ambiguous cases can result in the misidentification of
the host.
Other supernova surveys (e.g.\ SNLS) 
with spectroscopically confirmed host associations (e.g.\ with
matched transient and galaxy redshifts) that replicate the DES population
can be used to put a prior on this probability.

%An effective way of addressing this term is to
%analyze the supernovae assuming a correct host match, and
%experimentally determine $P(\text{mismatch} )$ and
%$\left. \langle \hat{\mu}-\mu(z)\rangle \right|_{\text{mismatch}}$, $\left. \langle \hat{z}-z \rangle \right|_{\text{mismatch}}$ and cross-terms to correct bias in the Hubble diagram.

% only model $T=\text{SN~Ia}$,
%and experimentally determine $P(\text{non-Ia})$, 
%$\left. \langle \mu-\mu_{model} \rangle \right|_{non-Ia}$, and its uncertainty for the
%sample of non-Ia's that are discovered and misclassified as SN~Ia.
%This measurement is achieved through spectroscopic
%typing of a subset drawn from the populations used in the Hubble diagram.
%The weighted bias would be added as a correction to the Hubble diagram under the assumption that all objects are SNe~Ia.

\subsection{Type~Ia Supernova Model}
\label{snmodel:sec}
The DES light curve analysis in principle may be used for typing and determining
redshift, in addition to getting supernova properties.   
Generally the relevant term is
\begin{equation}
P(\hat{D}_i=1,\hat{\mathit{ADU}}_{i}, \hat{T}_{Si}, \hat{z}_{Si}, \hat{\theta}_{Si}, \hat{z}_{Hi}, \hat{\theta}_{Hi}
 | \mu_i,z_i,T_i,\theta_T,\theta_{Ti},\phi(\lambda)).
 \end{equation}
Detection and typing depend on the realized light curves, and hence the truncated
distribution of discovered transients typed as SN~Ia.

The model contains a piece $P(\hat{\mathit{ADU}}_{i} \hat{z}_{Si}| \mu_i,z_i,\theta_T, \theta_{Ti}, ,\phi(\lambda))$, which
is the piece that describes underlying SN~Ia properties and light-curve fits.

\subsection{Detection Efficiency}
There term $P(\hat{D}|\mathit{ADU}_V)$.  For all high-redshift supernova searches
I know of, this depends on the signal-to-noise of the data, and very weakly on the
structure of the underlying galaxy.  Our detection is now based on machine learning,
we should confirm that detection dependencies do not depend on our model parameters
(except for RA/Dec).

\section{Systematic Tests}
\label{systematics:sec}
The Hubble diagrams of subsets of data quantify systematic uncertainty.  
Subsets that may be expected to show evidence for systematics include:
spectroscopically typed SNe~Ia; transients in deep (3) versus shallow (1/2) DES fields;
transients with NIR data; splits based on host properties.
